---
title: "Bike Sharing Analysis"
output: pdf_document
always_allow_html: yes
author: "Mohamad Makkawi"
---

```{r,warning=FALSE}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(leaflet))
suppressPackageStartupMessages(library(scales))
suppressPackageStartupMessages(library(kableExtra))
```

# Stations

```{r, echo=TRUE,warning=FALSE,message=FALSE}
stations <- read_csv("data/station_data.csv")
```

```{r}
summary(stations)
```

### Cleaning 'Stations' data

#### Duplicate station names

```{r}
head(stations %>%
  group_by(Name)%>%
  summarize(n_occurences=n()) %>%
  arrange(desc(n_occurences)))
```

Note: Civic Center BART, CalTrain 2, San Mateo County Center, Spear at Folsom are duplicates.

#### Duplicate station IDs

```{r}
stations[duplicated(stations$Name), ]
```

Note: The IDs to remove/combine from the trips dataset later are 85, 87, 88, 89, and 90.

#### Creating clean stations data

```{r}
station <- stations[!duplicated(stations$Name), ]

head(station)
```

### Number of stations per city

```{r}

station %>%
  group_by(City) %>%
  summarize(n_stations=n()) %>%
  ggplot(aes(City,n_stations,fill=City)) +
  geom_col() +
  ylab("Number of stations")+
  theme(legend.position = "none")
  

```

### Number of stations by dock count

```{r}
station %>%
  group_by(`Dock Count`) %>%
  summarize(`Number of Stations`=n())
```

### Visual distribution of the stations

#### Using leaflet

```{r}

factpal <- colorFactor(palette = "Set1",stations$City)

leaflet(options = leafletOptions(preferCanvas = TRUE)) %>% setView(lat = 37.55,lng = -122.2,zoom=10) %>% addProviderTiles(providers$Esri.WorldTopoMap) %>% addCircles(data=stations,~Long,~Lat,color=~factpal(City))
```

## Closer look at San Francisco

```{r}

leaflet(options = leafletOptions(preferCanvas = TRUE)) %>% setView(lat = 37.79,lng = -122.40,zoom=14) %>% 
  #addTiles() %>% 
  addProviderTiles(providers$Hydda.RoadsAndLabels) %>% 
  addProviderTiles(providers$Hydda.Base,options = providerTileOptions(opacity = 0.3))%>%
  #addProviderTiles(providers$Esri.NatGeoWorldMap,options = providerTileOptions(opacity = 0.35))%>%
  addCircleMarkers(data=stations,~Long,~Lat,color="Red",label = ~Name,radius = ~`Dock Count`/2.5,stroke = FALSE,fillOpacity = 0.5)
```


# Trip Data

```{r}

trips <- read_csv("data/trip_data.csv")

```

```{r}
head(trips)
```

```{r}
summary(trips)
```


### Cleaning 'Trips' Data

```{r}
trips_clean <- trips %>%
  rename(start_date = `Start Date`,start_station = `Start Station`,end_date = `End Date`,end_station = `End Station`,trip_id = `Trip ID`) %>%
  mutate(start_station = replace(start_station, start_station == 87, 49),
         start_station = replace(start_station, start_station == 88, 69),
         start_station = replace(start_station, start_station == 89, 72),
         start_station = replace(start_station, start_station == 90, 72),
         end_station = replace(end_station,end_station == 87, 49),
         end_station = replace(end_station,end_station == 88, 69),
         end_station = replace(end_station,end_station == 89, 72),
         end_station = replace(end_station,end_station == 90, 72))
```

### Combining Trips and Stations data

```{r}
trips_details <- trips_clean %>%
  left_join(stations,by=c("start_station"="Id")) %>%
  rename(start_city=City, start_name = Name, start_lat = Lat, start_long = Long, start_dock = `Dock Count`) %>%
  left_join(stations,by=c("end_station"="Id")) %>%
  rename(end_city = City,end_name = Name,end_lat = Lat,end_long = Long, end_dock = `Dock Count`) %>%
  select(trip_id,start_date,start_station,start_name,start_city,start_lat,start_long,start_dock,end_date,end_station,end_name,end_city,end_lat,end_long,end_dock,`Subscriber Type`) 

```

### Start and End Cities

```{r}
trips_details %>%
  group_by(start_city,end_city)%>%
  summarize(number_of_trips=n())%>%
  arrange(desc(number_of_trips))
```

Note: Almost all trips start and end in the same city. The vast majority of those being within the city of San Francisco.

## Trips within San Francisco

Seeing as the vast majority of trips in the trips dataset occurred within San Francisco city, we'll focus on these for the remainder of the analysis and building of the training set.

```{r}
sf_trips <- trips_details %>%
  filter(start_city == "San Francisco" & end_city == "San Francisco") %>%
  select(-start_city,-end_city)
```

```{r}
head(sf_trips)
```


### Visualizing volume of trips in San Francisco

Looking at most active stations for trips started (red) and trips ended (blue)

```{r}
sf_start <- sf_trips %>%
  group_by(start_station,start_lat,start_long,start_name)%>%
  summarize(n=n())
```

```{r}
sf_end <- sf_trips %>%
  group_by(end_station,end_lat,end_long,end_name)%>%
  summarize(n=n())
```

```{r}
leaflet(options = leafletOptions(preferCanvas = TRUE)) %>% setView(lat = 37.79,lng = -122.4,zoom=14) %>% 
  addProviderTiles(providers$Hydda.RoadsAndLabels) %>% 
  addProviderTiles(providers$Hydda.Base,options = providerTileOptions(opacity = 0.3))%>%
  addCircleMarkers(data=sf_start,~start_long,~start_lat,color="Red",group = "Start",popup=~ paste(
          start_name,
          "</br><b>Trips Started:</b>",
          n
        ),radius = ~n/1200,stroke = FALSE,fillOpacity = 0.5)%>%
  addCircleMarkers(data=sf_end,~end_long,~end_lat,color="Blue",group="End",popup=~ paste(
          end_name,
          "</br><b>Trips Ended:</b>",
          n
        ),radius = ~n/1200,stroke=FALSE,fillOpacity = 0.5) %>%
  addLayersControl(
    overlayGroups = c("Start", "End"),
    options = layersControlOptions(collapsed = FALSE)
  )
```

### Most popular starting stations

```{r}
## Refactoring start_name and end_name to help with sorting

sf_trips$start_name <- factor(sf_trips$start_name, levels=unique(as.character(sf_trips$start_name)) )
sf_trips$end_name <- factor(sf_trips$end_name, levels=unique(as.character(sf_trips$end_name)) )

```

```{r}

sf_trips %>%
  group_by(start_station,start_name) %>%
  summarize(n=n())%>%
  ggplot(aes(x=reorder(start_name,n),n))+
  geom_col()+
  theme(axis.text.x = element_text(size=7,angle = 45, hjust = 1))+
  ylab("Number of trips started")+xlab("")+
  ggtitle("Number of trips started by station name")
```

### Most popular ending stations

```{r}

sf_trips %>%
  group_by(end_station,end_name) %>%
  summarize(n=n())%>%
  ggplot(aes(x=reorder(end_name,n),n))+
  geom_col()+
  theme(axis.text.x = element_text(size=7,angle = 45, hjust = 1))+
  ylab("Number of trips ended")+xlab("")+
  ggtitle("Number of trips ended by station name")


```

### Side-by-side comparison

```{r}
### Getting dataframe into the format that allows 'dodge2' visualization

sf_start<-sf_trips %>%
  group_by(start_station,start_name) %>%
  summarize(n_start=n())%>%
  mutate(type = "Start")

sf_end<-sf_trips %>%
  group_by(end_station,end_name) %>%
  summarize(n_end=n()) %>%
  mutate(type = "End") %>%
  rename(start_station = end_station,start_name = end_name,n_start=n_end)

viz_sf <- rbind(sf_start,sf_end)

```


```{r}
### Refactoring to show start before end

viz_sf$type <- factor(viz_sf$type, levels= c("Start","End"))

```


```{r}
viz_sf %>%
  ggplot(aes(reorder(start_name,n_start),n_start,fill=type))+
  geom_col(position="dodge2")+
  theme(axis.text.x = element_text(size=7,angle = 45, hjust = 1),legend.title = element_blank(),legend.position = c(.05, .95),
    legend.justification = c("left", "top"),
    legend.box.just = "right",
    legend.margin = margin(6, 6, 6, 6))+
  ylab("Number of trips")+xlab("")+
  ggtitle("Number of trips started/ended by station name")
           
```

-----------------------


```{r}
head(trips)%>%
  kable() %>%
  kable_styling(bootstrap_options = "basic")
```

# Number of trips started by station

```{r}
trips %>%
  group_by(start_station_id = `Start Station`) %>%
  summarize(n=n()) %>%
  arrange(desc(n)) %>%
  ggplot(aes(start_station_id,n)) +
  geom_col()
```

# Number of trips started by city

```{r}
stations_trips <- stations %>%
  select(Id,City)
```


```{r}
trips %>%
  select(everything(),-`End Date`,-`End Station`) %>%
  left_join(stations_trips,by = c("Start Station"="Id")) %>%
  group_by(City) %>%
  summarize(n=n())

#%>%
 # kable() %>%
  #kable_styling(bootstrap_options = "basic")
```

### Takeaway: Large majority of trips started from San Francisco

# Number of trips by end station

```{r}
trips %>%
  group_by(end_station_id = `End Station`) %>%
  summarize(n=n()) %>%
  ggplot(aes(end_station_id,n)) +
  geom_col()
```

## Number of trips ended by City

```{r}
trips %>%
  select(everything(),-`Start Date`,-`Start Station`) %>%
  left_join(stations_trips,by = c("End Station"="Id")) %>%
  group_by(City) %>%
  summarize(n=n())

#%>%
 # kable() %>%
  #kable_styling(bootstrap_options = "basic")
```

### Takeaway: Most trips start and end in the same city

## Most popular routes

```{r}
head(pop_routes <- trips %>%
  group_by(start_id = `Start Station`,end_id = `End Station`)%>%
  summarize(n=n()) %>%
  arrange(desc(n)),10)

#%>%
  #kable() %>%
  #kable_styling(bootstrap_options = "basic")
```

### Visualizing popular routes in San Francisco

```{r}
stations_coor <- stations %>%
  select(Id,Lat,Long)
```

```{r}
pop_routes %>%
  filter(start_id > 40 & start_id < 75 & end_id>40 & end_id <75) %>%
  left_join(stations_coor, by = c("start_id"="Id")) %>%
  select(start_id,Lat1 = Lat,Long1 = Long,end_id,n) %>%
  left_join(stations_coor,by=c("end_id"="Id")) %>%
  ggplot()+
  geom_segment(aes(x=Long1,y=Lat1,xend=Long,yend=Lat,color=n/1000))
```

## Trip durations

```{r}
tim_dif <- trips %>%
  mutate(start = lubridate::dmy_hm(`Start Date`),end = lubridate::dmy_hm(`End Date`)) %>%
  mutate(time_diff = end-start) %>%
  select(`Trip ID`,`Start Station`,start,`End Station`,end,time_diff)
```

```{r}
head(tim_dif)%>%
  kable() %>%
  kable_styling(bootstrap_options = "basic")
```

```{r}

tim_dif %>%
  mutate(duration = if_else(time_diff>1440,"Long",if_else(time_diff<60,"Short","Average"))) %>%
  group_by(duration) %>%
  summarize(n=n())%>%
  kable() %>%
  kable_styling(bootstrap_options = "basic")

```

### Takeaway: Vast majority of trips less than 1 hour.

## Analysis within short trips

Note: Short trips defined as trips less than 60 minutes

```{r}
tim_dif %>%
  filter(time_diff<60) %>%
  mutate(duration = if_else(time_diff>45,"45+",if_else(time_diff>30,"30-45",if_else(time_diff>15,"15-30","0-15")))) %>%
  group_by(duration) %>%
  summarize(n=n()) %>%
  ggplot(aes(duration,n))+
  scale_y_continuous(breaks = pretty_breaks())+
  geom_col() + xlab("Duration (in minutes)") + ylab("Number of trips") + ggtitle("Breakdown of trip durations less than 60 minutes")
```

### Takeaway: Most bike trips are less than 15 minutes in duration

## Time of day

```{r}
trips %>%
  mutate(start = lubridate::dmy_hm(`Start Date`),end = lubridate::dmy_hm(`End Date`)) %>%
  select(`Trip ID`,`Start Station`,start,`End Station`,end) %>%
  group_by(time_of_day = hour(start)) %>%
  summarize(n=n()) %>%
  ggplot(aes(time_of_day,n)) +
  geom_col() +
  xlab("Time of day") + ylab("Number of bikes rented")
```

### Takeaway: Most bike trips occur between 7AM and 7PM. There is a morning rush (7AM-9AM) and an evening rush (4PM-6PM).

## Time of year

```{r}
trips %>%
  mutate(start = lubridate::dmy_hm(`Start Date`),end = lubridate::dmy_hm(`End Date`)) %>%
  select(`Trip ID`,`Start Station`,start,`End Station`,end) %>%
  mutate(my_date = as.Date(start))%>%
  group_by(m_date = month(my_date)) %>%
  summarize(n=n()) %>%
  ggplot(aes(m_date,n))+
  geom_col(bins=10)+
  scale_x_continuous(breaks= pretty_breaks())+
  xlab("Month")+ylab("Number of bikes rented")
```

### Takeaway: Noticeable dip in bike trips during Winter months

## Day of week

```{r}
trips %>%
  mutate(start = lubridate::dmy_hm(`Start Date`),end = lubridate::dmy_hm(`End Date`)) %>%
  select(`Trip ID`,`Start Station`,start,`End Station`,end) %>%
  mutate(my_date = as.Date(start)) %>%
  mutate(day_of_week = wday(my_date)) %>%
  group_by(day_of_week) %>%
  summarize(n=n()) %>%
  ggplot(aes(day_of_week,n))+
  geom_col() +
  xlab("Day of the week") + ylab("Number of bikes rented")
```

### Takeaway: Significant amount of trips occur on weekdays

# Finding the net pickups/dropoffs for each station in a given hour

Converting to datetime object

```{r}
trips_dt <- trips %>%
  mutate(start = lubridate::dmy_hm(`Start Date`),end = lubridate::dmy_hm(`End Date`)) %>%
  select(`Trip ID`,`Start Station`,start,`End Station`,end)
```

```{r}
head(trips_dt)%>%
  kable() %>%
  kable_styling(bootstrap_options = "basic")
```

Breaking down date variable

```{r}
trips_d <- trips_dt %>%
  mutate(start_year = year(start),start_month = month(start),start_day = day(start),start_hour = hour(start),end_year = year(end),end_month = month(end),end_day = day(end),end_hour = hour(end)) %>%
  select(`Trip ID`,start_year,start_month,start_day,start_hour,`Start Station`,end_year,end_month,end_day,end_hour,`End Station`)
```

```{r}
head(trips_d)%>%
  kable() %>%
  kable_styling(bootstrap_options = "basic")
```

## Trips started from specific station in a given hour

```{r}
trips_start <- trips_d %>%
  group_by(start_year,start_month,start_day,start_hour,`Start Station`) %>%
  summarize(n_start=n()) %>%
  arrange(desc(n_start))
```

```{r}

head(trips_start)%>%
  kable() %>%
  kable_styling(bootstrap_options = "basic")

```

## Trips ended from specific station in a given hour

```{r}
trips_end <- trips_d %>%
  group_by(end_year,end_month,end_day,end_hour,`End Station`) %>%
  summarize(n_end=n()) %>%
  arrange(desc(n_end))
```

```{r}
head(trips_end)%>%
  kable() %>%
  kable_styling(bootstrap_options = "basic")
```

Combining the two tables

```{r}

trips_diff <- trips_start %>%
  left_join(trips_end, by = c("start_year"="end_year","start_month"="end_month","start_day"="end_day","start_hour"="end_hour"))

```

#### Net change in bikes in a given hour

```{r}

trips_diff$date <- paste(trips_diff$start_day, trips_diff$start_month, trips_diff$start_year, sep="/") %>% dmy() %>% as.Date()

head(trips_df <- trips_diff %>%
  filter(`End Station`==`Start Station`) %>%
  mutate(net = n_end - n_start,station = `Start Station`,hour = start_hour) %>%
  select(date,hour,station,net)) %>%
  kable() %>%
  kable_styling(bootstrap_options = "basic")
```

### Example: Fluctuation of net bikes at Station 70 on a selected day

```{r}

trips_df %>%
  filter(station == 70 & date == "2015-01-28") %>%
  ggplot(aes(start_hour,net)) +
  geom_line() +
  xlab("Hour of day") + ylab("Net change in bikes") + ggtitle("Fluctuation at Station 70 on a select day")

```

This appears to be a neighborhood where individuals commute out to go to work every morning by bike, and return in the evening - hence, the fluctuating net pickups/dropoffs.

## Creating the training set

```{r}
t_set <- trips_df %>%
  ungroup() %>%
  select(date,hour,station,net) #%>%
  #write.csv("training_set.csv")
```


```{r}
stations_city <- stations %>%
  select(Id,City)
```


```{r}
t_set_2 <- t_set %>%
  left_join(stations_city,by=c("station"="Id"))
```


# Weather dataset

```{r}

weather <- read_csv("data/weather_data.csv")

```

```{r}
weather_df<-weather %>%
  mutate(date_format = lubridate::dmy(Date)) %>%
  mutate(events = if_else(is.na(Events),"None",Events)) %>%
  select(date_format,temp = `Mean TemperatureF`,Zip,events,PrecipitationIn,wind_speed = `Mean Wind SpeedMPH`) %>%
  mutate(City = if_else(Zip==94107,"San Francisco",if_else(Zip==94063,"Redwood City",if_else(Zip==94301,"Palo Alto",if_else(Zip==94041,"Mountain View","San Jose"))))) %>%
  select(everything(),-Zip)
```

```{r}
weather_df <- weather_df %>%
  mutate(is_raining = if_else(events == "Rain" | events == "Fog-Rain",1,0))
```

```{r}
head(weather_df)
```

### Joining weather and net change tables

```{r}

t_set_3 <- t_set_2 %>%
  left_join(weather_df,by = c("date" = "date_format","City"="City"))

```

```{r}
head(t_set_3)%>%
  kable() %>%
  kable_styling(bootstrap_options = "basic")
```


### Creating is_weekday and is_winter features and outputting training set

```{r}
q<-t_set_3 %>%
  mutate(day_week = wday(date)) %>%
  mutate(is_weekday = if_else(day_week %in% c(2,3,4,5,6),1,0)) %>%
  mutate(is_winter = if_else(month(date) %in% c(11,12,1,2),1,0)) %>%
  mutate(time_of_day = if_else(hour %in% c(7,8,9),"Morning",if_else(hour %in% c(10,11,12,13,14,15),"Midday",if_else(hour %in% c(16,17,18),"Afternoon",if_else(hour %in% c(19,20,21,22,23),"Night","Overnight"))))) %>%
  select(everything(),-date,-day_week,-events) %>%
  filter(!is.na(temp)) %>%
  write.csv("training_set_2.csv")

```

Check out the Jupyter Notebook titled `Bike_Modelling.ipynb` for the modelling section.

